---
title: "DownClim - Downscale Climate Projections"
author: Sylvain Schmitt -
date: today
date-format: medium
format: 
  gfm:
    toc: true
    toc-depth: 2
---

```{r setup, include=FALSE}
rm(list = ls()) ; invisible(gc()) ; set.seed(42)
library(knitr)
library(tidyverse)
theme_set(bayesplot::theme_default())
opts_chunk$set(echo = F, message = F, warning = F, fig.height = 6, fig.width = 8,
               cache = F, cache.lazy = F)
```

[`snakemake` & `singularity`](https://github.com/sylvainschmitt/snakemake_singularity) workflow to downscale climate projections

**Description.**

```{r dag, fig.cap="Workflow."}
knitr::include_graphics("dag/dag.svg")
```

# Installation

This workflow is built on:

-   [x] Python ≥3.5
-   [x] Snakemake ≥5.24.1
-   [x] Singularity ≥3.7.3

Once installed simply clone the workflow:

```{bash, eval=F, echo=T}
git clone git@github.com:sylvainschmitt/DownClim.git
cd DownClim
```

# Credentials

*Detail how to get the different credentials.*

# Usage

```{bash, eval=F, echo=T}
module load bioinfo/Snakemake/7.20.0 # for test on nod depending on your HPC
snakemake -np # dry run
snakemake --dag | dot -Tsvg > dag/dag.svg # dag
snakemake -j 1 --resources mem_mb=10000 # local run (test)
sbatch job_muse.sh # HPC run with slurm
```

# Workflow

### [rule](https://github.com/sylvainschmitt/smkTemplate/blob/main/rules/rule.smk)

-   Script: [`rule.R`](https://github.com/sylvainschmitt/smkTemplate/blob/main/scripts/rule.R)

Dummy example of a rule based on an R script.

Alternative to get CORDEX:

-   IPSL: https://esgf-node.ipsl.upmc.fr/search/cordex-ipsl/
-   CDS: https://cds.climate.copernicus.eu/cdsapp#!/dataset/projections-cordex-domains-single-levels?tab=overview
-   esgf-pylient https://esgf-pyclient.readthedocs.io/en/latest/notebooks/demo/subset-cmip5.html
-   https://github.com/ESGF/sproket
-   https://pypi.org/project/esgf-scraper/
-   https://docs.esmvaltool.org/projects/ESMValCore/en/latest/api/esmvalcore.esgf.html

# Data

*Detail the different data sources.*
